---
title: "Validation of flipmeta"
format: html
execute:
  warning: false
  message: false
---

```{r}
library(metafor)
library(flipscores)

devtools::load_all()

set.seed(2026)
options(scipen = 999)

dat <- escalc(
    measure = "RR",
    ai = tpos,
    bi = tneg,
    ci = cpos,
    di = cneg,
    data = dat.bcg
)

dat$vi_equal <- 0.01

flips <- flipscores:::make_flips(
    nrow(dat),
    n_flips = 1000
)
```

# Equal sampling variances

When all sampling variances are equal and $\tau^2 = 0$, the meta-analytic weight matrix is proportional to the identity matrix. In this case, `flipscores` provides a useful benchmark for the implementation in `flipmeta`.

## Intercept-only model

```{r}
fit_glm <- glm(
    yi ~ 1,
    data = dat
)

fit_rma <- rma(
    yi,
    vi_equal,
    data = dat,
    method = "EE"
)

fit_fs <- flipscores(
    fit_glm,
    flips = flips
)

fit_fm <- flipmeta(
    fit_rma,
    flips = flips
)

fit_fs$summary_table
fit_fm$summary_table
```

The two permutation spaces should differ only by a constant scale factor.

```{r}
ratio <- fit_fm$Tspace[, 1] / fit_fs$Tspace[, 1]

range(ratio)
sd(ratio)

c(
    flipscores = fit_fs$summary_table$p,
    flipmeta = fit_fm$summary_table$p
)
```

For the intercept-only model, the difference in scale can also be obtained from the empirical scale used by `flipscores` relative to the known meta-analytic sampling standard deviation.

```{r}
sqrt(mean(dat$yi^2)) / sqrt(dat$vi_equal[1])
```

# One moderator

```{r}
fit_glm <- glm(
    yi ~ ablat,
    data = dat
)

fit_rma <- rma(
    yi,
    vi_equal,
    mods = ~ ablat,
    data = dat,
    method = "EE"
)

fit_fs <- flipscores(
    fit_glm,
    flips = flips
)

fit_fm <- flipmeta(
    fit_rma,
    flips = flips
)

fit_fs$summary_table
fit_fm$summary_table
```

The proportionality is checked separately for the intercept and the moderator because each coefficient is tested under a different null model.

```{r}
ratio_intercept <- fit_fm$Tspace[, "intrcpt"] /
    fit_fs$Tspace[, "(Intercept)"]

ratio_ablat <- fit_fm$Tspace[, "ablat"] /
    fit_fs$Tspace[, "ablat"]

c(
    intercept_min = min(ratio_intercept),
    intercept_max = max(ratio_intercept),
    intercept_sd = sd(ratio_intercept)
)

c(
    ablat_min = min(ratio_ablat),
    ablat_max = max(ratio_ablat),
    ablat_sd = sd(ratio_ablat)
)

cbind(
    flipscores = fit_fs$summary_table$p,
    flipmeta = fit_fm$summary_table$p
)
```

# Two moderators

We now test the same equivalence when residualization with respect to nuisance covariates is required. For the test of `ablat`, the nuisance design matrix contains the intercept and `year`.

```{r}
fit_glm <- glm(
    yi ~ ablat + year,
    data = dat
)

fit_rma <- rma(
    yi,
    vi_equal,
    mods = ~ ablat + year,
    data = dat,
    method = "EE"
)

fit_fs <- flipscores(
    fit_glm,
    flips = flips
)

fit_fm <- flipmeta(
    fit_rma,
    flips = flips
)

fit_fs$summary_table
fit_fm$summary_table
```

```{r}
ratio_ablat <- fit_fm$Tspace[, "ablat"] /
    fit_fs$Tspace[, "ablat"]

range(ratio_ablat)
sd(ratio_ablat)

cbind(
    flipscores = fit_fs$summary_table$p,
    flipmeta = fit_fm$summary_table$p
)
```

# Heterogeneous sampling variances

With heterogeneous sampling variances and $\tau^2 = 0$,

$$
\operatorname{Var}(y_i) = v_i
$$

and therefore

$$
W = \operatorname{diag}(1/v_i).
$$

A weighted Gaussian GLM does not provide the same direct benchmark for the sign-flipping standardization. We therefore compare `flipmeta` directly with the analytical expressions obtained by specializing the score sign-flipping test to the meta-regression model.

We consider `ablat` as the focal moderator and `year` as a nuisance moderator.

```{r}
fit_rma <- rma(
    yi,
    vi,
    mods = ~ ablat + year,
    data = dat,
    method = "EE"
)

fit_fm <- flipmeta(
    fit_rma,
    flips = flips
)

fit_fm$summary_table
```

## Score contributions

Let $x$ denote the focal moderator and $Z$ the nuisance design matrix. Under $H_0:\beta_x=0$,

$$
\hat\gamma_0 =
(Z^\top WZ)^{-1} Z^\top Wy
$$

and

$$
r_0 = y-Z\hat\gamma_0.
$$

The focal predictor is residualized with respect to the nuisance covariates as

$$
\tilde{x}
=
x-Z(Z^\top WZ)^{-1}Z^\top Wx.
$$

The individual score contributions are then

$$
\nu_i = \tilde{x}_i w_i r_{0i}.
$$

```{r}
y <- dat$yi
x <- dat$ablat
Z <- model.matrix(~ year, data = dat)

k <- length(y)

w <- 1 / dat$vi
W <- diag(w)

gamma0 <- solve(
    crossprod(Z, W %*% Z),
    crossprod(Z, W %*% y)
)

r0 <- drop(
    y - Z %*% gamma0
)

gamma_x <- solve(
    crossprod(Z, W %*% Z),
    crossprod(Z, W %*% x)
)

x_tilde <- drop(
    x - Z %*% gamma_x
)

score_contributions <- x_tilde * w * r0

score_oracle <- sum(score_contributions)

score_flipmeta <- fit_fm$summary_table$score[
    fit_fm$summary_table$coefficient == "ablat"
]

c(
    oracle = score_oracle,
    flipmeta = score_flipmeta,
    ratio = score_flipmeta / score_oracle
)
```

## Standardized flipped statistics

Define

$$
H =
W^{1/2}Z
(Z^\top WZ)^{-1}
Z^\top W^{1/2}
$$

and

$$
P=I-H.
$$

For flip $b$, the flipped score is

$$
S_b = \sum_i f_{bi}\nu_i,
$$

and the flip-specific variance term is

$$
C_b =
x^\top
W^{1/2}
P F_b P F_b P
W^{1/2}x.
$$

The standardized statistic is

$$
T_b = \frac{S_b}{\sqrt{C_b}}.
$$

```{r}
sqrtW <- diag(sqrt(w))

H <- sqrtW %*%
    Z %*%
    solve(crossprod(Z, W %*% Z)) %*%
    t(Z) %*%
    sqrtW

P <- diag(k) - H

score_flipped <- drop(
    flips %*% score_contributions
)

variance_flipped <- numeric(nrow(flips))

for (b in seq_len(nrow(flips))) {

    F <- diag(flips[b, ])

    variance_flipped[b] <- drop(
        t(x) %*%
            sqrtW %*%
            P %*%
            F %*%
            P %*%
            F %*%
            P %*%
            sqrtW %*%
            x
    )
}

T_oracle <- score_flipped / sqrt(variance_flipped)
T_flipmeta <- fit_fm$Tspace[, "ablat"]

ratio <- T_flipmeta / T_oracle

range(ratio)
sd(ratio)
```

```{r}
p_oracle <- mean(
    abs(T_oracle) >= abs(T_oracle[1])
)

p_flipmeta <- fit_fm$summary_table$p[
    fit_fm$summary_table$coefficient == "ablat"
]

c(
    oracle = p_oracle,
    flipmeta = p_flipmeta
)
```

## Numerical checks

```{r}
crossprod(Z, W %*% r0)

crossprod(Z, W %*% x_tilde)

c(
    H_symmetric = max(abs(H - t(H))),
    H_idempotent = max(abs(H %*% H - H)),
    P_idempotent = max(abs(P %*% P - P))
)
```

# Heterogeneity estimated under the null

For a random-effects meta-regression, $\tau^2$ is treated as a nuisance parameter and is estimated under the coefficient-specific null hypothesis.

For the test of `ablat`,

$$
H_0:\beta_{\mathrm{ablat}}=0,
$$

so that the null model contains the intercept and `year` only.

```{r}
fit_rma <- rma(
    yi,
    vi,
    mods = ~ ablat + year,
    data = dat,
    method = "REML"
)

fit_fm <- flipmeta(
    fit_rma,
    flips = flips
)

fit_fm$summary_table
```

## Null heterogeneity estimate

The null model is fitted explicitly to obtain the oracle value of $\hat\tau_0^2$.

```{r}
fit_null <- rma(
    yi,
    vi,
    mods = ~ year,
    data = dat,
    method = "REML"
)

tau2_null <- fit_null$tau2

c(
    oracle = tau2_null,
    flipmeta = fit_fm$summary_table$tau2_null[
        fit_fm$summary_table$coefficient == "ablat"
    ]
)
```

## Score contributions

Conditional on $\hat\tau_0^2$, the weights are

$$
\hat w_{0i}
=
\frac{1}{v_i+\hat\tau_0^2}.
$$

```{r}
y <- dat$yi
x <- dat$ablat
Z <- model.matrix(~ year, data = dat)

k <- length(y)

w <- 1 / (dat$vi + tau2_null)
W <- diag(w)

gamma0 <- solve(
    crossprod(Z, W %*% Z),
    crossprod(Z, W %*% y)
)

r0 <- drop(
    y - Z %*% gamma0
)

gamma_x <- solve(
    crossprod(Z, W %*% Z),
    crossprod(Z, W %*% x)
)

x_tilde <- drop(
    x - Z %*% gamma_x
)

score_contributions <- x_tilde * w * r0

score_oracle <- sum(score_contributions)

score_flipmeta <- fit_fm$summary_table$score[
    fit_fm$summary_table$coefficient == "ablat"
]

c(
    oracle = score_oracle,
    flipmeta = score_flipmeta,
    ratio = score_flipmeta / score_oracle
)
```

## Standardized flipped statistics

```{r}
sqrtW <- diag(sqrt(w))

H <- sqrtW %*%
    Z %*%
    solve(crossprod(Z, W %*% Z)) %*%
    t(Z) %*%
    sqrtW

P <- diag(k) - H

score_flipped <- drop(
    flips %*% score_contributions
)

variance_flipped <- numeric(nrow(flips))

for (b in seq_len(nrow(flips))) {

    F <- diag(flips[b, ])

    variance_flipped[b] <- drop(
        t(x) %*%
            sqrtW %*%
            P %*%
            F %*%
            P %*%
            F %*%
            P %*%
            sqrtW %*%
            x
    )
}

T_oracle <- score_flipped / sqrt(variance_flipped)
T_flipmeta <- fit_fm$Tspace[, "ablat"]

ratio <- T_flipmeta / T_oracle

range(ratio)
sd(ratio)
```

```{r}
p_oracle <- mean(
    abs(T_oracle) >= abs(T_oracle[1])
)

p_flipmeta <- fit_fm$summary_table$p[
    fit_fm$summary_table$coefficient == "ablat"
]

c(
    oracle = p_oracle,
    flipmeta = p_flipmeta
)
```

## Numerical checks

```{r}
crossprod(Z, W %*% r0)

crossprod(Z, W %*% x_tilde)

c(
    H_symmetric = max(abs(H - t(H))),
    H_idempotent = max(abs(H %*% H - H)),
    P_idempotent = max(abs(P %*% P - P))
)
```
